### 2016-06-26 reasoning.tex
edits to secs. 1 - 4

We now describe a proof system that lets us demonstrate that a \txnimp
program satisfies its high-level invariants in all small-step
executions that satisfy a chosen trace invariant\footnote{It is
important to mind the distinction between high-level invariants
(\eg~\C{X$\ge$0}) and trace invariants (\eg~\C{RC}). The former
constitute \emph{proof} \emph{obligations} for the programmer, whereas
the latter capture \emph{assumptions} about the machine. } ($\I$).
Since the trace invariant captures transaction isolation constraints,
the demonstration is a proof that the given selection of isolation
levels for transactions is sufficient to enforce program's invariants.

As mentioned earlier, our proof system is a rely-guarantee framework
that admits compositional reasoning by abstracting away the
environmental interference as a rely relation. A conspicuous
difference between a standard development of rely-guarantee and ours
is that, while the former reasons in terms of program states (variable
to value bindings), we reason in terms of executions as captured by
their traces ($\E$). In particular, our rely ($R$) and guarantee ($G$)
relations relate executions (i.e., $R,\,Q\,\subseteq\,\E\times\E$),
and pre ($P$) and post ($Q$) conditions are assertions over executions
(i.e., $P,\,Q\, : \E \rightarrow \Prop$). Our development nonetheless
facilitates state-based reasoning via the interpretation function
($\interp{\cdot}$) introduced in \S\ref{sec:syntax}. For example, if a
\emph{bi-state} rely relation relates every pair of states $\sigma$
and $\sigma'$ such that $\sigma'(X) \ge \sigma(X) \ge 0$, the
corresponding \emph{bi-execution} rely relation relates every pair of
executions $\E$ and $\E'$ such that $\interp{\E}(X) \ge
\interp{\E'}(X) \ge 0$.  Assertions on state are also written
similarly. For instance, consider the postcondition
$\C{C\,=\,k-a1-a2}$ of the program in Fig.~\ref{fig:motiv-eg-1}. The
corresponding assertion on the post-state
($\lambda\sigma.~\sigma(\C{C}) \,=\, \C{k-a1-a2}$) asserts that in all
states resulting from executing the program, the value of \C{C} is
\C{k-a1-a2}. The equivalent execution-based assertion
($\lambda\E.~\interp{\E}(\C{C})\,=\,\C{k-a1-a2}$) asserts that in all
executions of the program, the value of \C{C} is \C{k-a1-a2}.
However, having access to an execution facilitates assertions that go
beyond the state. An example is an invariant of \C{Wd1} described
abstractly in \S\ref{sec:motivation}, and reified as an
execution-based assertion below:
\begin{smathpar}
\begin{array}{l}
  \lambda\E.~ \neg(\underE{\committed(\C{Wd1}}) \conj 
      \underE{\C{Wd1} \wrstoar \C{C}}\\
  \hspace*{0.2in}\conj(\neg(\underE{\committed(\C{Wd2})}) \Rightarrow 
      \C{\interp{\E}(C) = k-a1}) \\
   \hspace*{0.2in}\conj (\underE{\committed(\C{Wd2})} \Rightarrow 
      \C{\interp{\E}(C) = k-a1-a2})
\end{array}
\end{smathpar}
% Where, $\committed$ is defined as:
% \begin{smathpar}
% \begin{array}{lcl}
% \underE{\committed(T_i)} & \Leftrightarrow &
%   \exists\eta.~\underE{\eta\in T_i} \conj \kind(\eta) = \C{COMMIT}\\
% \end{array}
% \end{smathpar}
One of the conjuncts asserts that \C{Wd1} has written to \C{C} - a
fact which cannot be deduced solely from the value of \C{C} (esp. in
presence of interference), but can be expressed as a proposition over
the execution trace.

@@

The standard rely guarantee judgment is the quintuple
$\rg{P,R}{c}{G,Q}$, which intuitively means the following: if a
command $c$ is executed in a state that satisfies its precondition
$P$, provided that every interference step during the execution is
contained inside the rely relation $R$, then the effect that each step
of executing $c$ has on the state is captured by $G$, and the final
state of execution satisfies the postcondition $Q$. The terms
``execution'' and ``step'' are defined with respect to an operational
semantics, which, in our case, is parameterized on the trace invariant
$\I$. Consequently, our rely-guarantee judgment for commands is the
sextuple $\rg{\I,P,R}{c}{G,Q}$, whose semantics differs from the
quintuple in that it requires every step of the execution to preserve
the trace invariant $\I$ for the postcondition to be valid.  As usual,
if $c$ is inside a transaction $T_i$, we write $\txnbox{c}_i$. \txnimp
expressions are side-effecting (they generate \C{RD} effects), and
admit interference during their evaluation.  Therefore, like the
commands, expressions need to be reasoned about explicitly. But,
unlike commands, expressions evaluate to values, and the reasoning
framework should also admit assertions on such values.  We therefore
define a separate judgment for expressions - the septuple
$\rg{\I,P,R}{e}{G,C,Q}$. The new entrant $C : \mathbb{N} \rightarrow
\Prop$ is an assertion on the return value of $e$.

The rules that define rely-guarantee judgment are shown in
Fig.~\ref{fig:rg-rules}. Like the standard definitions of rely
guarantee, our definition also relies on the \emph{stability}
condition, which requires pre and post conditions to hold despite the
interference from concurrent threads (captured by $R$). The stability
of pre and post conditions can be predicated on the assumption that
the interference preserves the trace invariant $\I$. Formally:
\begin{smathpar}
\begin{array}{lcl}
  \underI{\stable(R,P)} & \defeq & \forall \E, \E'.\, 
  \I(\E) \conj P(\E) \conj R(\E,\E') \\
  &   & \hspace*{1in}\conj \I(\E') \Rightarrow P(\E')\\
\end{array}
\end{smathpar}
However, the assumption that the interference preserves $\I$ (or,
dually, $\I$ withstands inteferference) needs to be justified
separately. We call this the stability requirement on $\I$:
\begin{smathpar}
\begin{array}{lcl}
\stable(R,\I)& \defeq & \forall \E, \E'.\, 
  \I(\E) \conj R(\E,\E')\Rightarrow \I(\E')\\
\end{array}
\end{smathpar}

The rule \rulelabel{RG-Var} defines the rely-guarantee judgment for
shared variable reads inside a transaction $T_i$. It requires $\I$ to
be stable, and pre and post conditions to be stable relative to $\I$.
The quantified premise effectively requires a proof that if the
abstract machine of Fig.~\ref{fig:txnimp} takes a step starting from
an execution $\E$ that satisfies the precondition $P$, then the
resultant execution $\E'$ satisfies the postcondition $Q$, and that
the guarantee $G$ faithfully captures the transition from $\E$ to
$\E'$. It is informative to compare this premise with the premise of
the \rulelabel{E-Aux} reduction rule of Fig.~\ref{fig:txnimp}. Similar
premises also appear in \rulelabel{RG-Asgn} and \rulelabel{RG-Txn}
rules, which define rely-guarantee judgments for assignments and
transactions, respectively. \rulelabel{RG-Var} however also requires
the return value ($\interp{S}(X)$) of the read to satisfy the
assertion $C$ meant for the value. \rulelabel{RG-Arith} defines the
RG judgment for an arithmetic expression $e_1\pm e_2$ in terms of the
corresponding judgments for the constitutent expressions $e_1$ and
$e_2$. The quantified premise requires any value resulting from
evaluating $e_1 \pm e_2$ to satisfy the assertion $C$, provided that
$e_1$ and $e_2$ always evaluate to values that satisfy $C_1$ and
$C_2$, respectively. The rules for sequential and parallel composition
of comands are more-or-less same as the corresponding rules in
classical rely-guarantee, hence elided.

@@

The \rulelabel{RG-Conseq} rule defines ways to strengthen or weaken
relations and assertions associated with the RG judgment of
transaction-bound expressions. Similar rules exist for
transaction-bound and top-level commands, but are not shown. In
particular, the rules allow the precondition $P$ and the rely relation
$R$ to be strengthened\footnote{A stronger relation is a smaller set
than a weaker relation, hence a stronger $R$ admits lesser
interference than a weaker $R$}, and the postcondition $Q$ (also, $C$
in the case of expressions) and the guarantee relation $G$ to be
weakened. This kind of strengthening/weakening is also admitted in the
classical RG framework. The most notable aspect of
\rulelabel{RG-Conseq} rules however is that they allow the trace
invariant $\I$ to be strengthened. Considering that $\I$ captures
isolation properties, this means that a program proved correct under
weaker isolation levels is also correct under stronger levels.
Parametricity over the trace invariant $\I$, combined with the ability
to strengthen $\I$ as needed, allows our proof system to admit a
highly flexible proof strategy to prove programs correct on various
stores. For example, programmers can \emph{define} isolation
guarantees of their choice (by defining $\I$ appropriately) and then
prove programs correct assuming the guarantees hold. Soundness of
strengthening $\I$ assures that the program can now be executed safely
on any real store that offers same or stronger isolation guarantees
than those assumed. The following subsection proves the soundness of
our proof system.

@@

We now formalize the semantics of the rely-guarantee judgment of
Fig.~\ref{fig:rg-rules} through a series of definitions.

\begin{definition}[\bfseries Step-indexed reflexive transitive closure]
Forall $A$, $R: A \rightarrow A \rightarrow \mathbb{P}$, and $n :
\mathbb{N}$, step-indexed reflexive transitive closure $R^n$ of $R$ is
the smallest relation satisfying the following
properties:
\begin{itemize}
\item $\forall (x:A).\, R^0 (x,x)$
\item $\forall (n:\mathbb{N})(x,y,z : A).\, R(x,y) \conj R^n(y,z) \Rightarrow
R^{n+1}(x,z)$
\end{itemize}
\end{definition}

\begin{definition}[\bfseries Interleaved step and multi-step relations]
Interleaved step relation interleaves the thread-local reduction with
the interference from concurrent threads captured as the rely relation
($R$).  It is defined as following:
\begin{smathpar}
\begin{array}{lcl}
\I \vdash (c,\E) \rstepsto (c',\E') & \defeq & \I \vdash 
  (c,\E) \stepsto (c',\E') \\
  &   & \disj (c' = c \conj R(\E, \E') \conj \I(\E'))\\
\end{array}
\end{smathpar}
The interleaved step relation for transaction bound expressions
($\txnbox{e}_i$) and commands ($\txnbox{c}_i$) is defined similarly.
Interleaved multi-step relation ($\stepssto{n}$) is the step-indexed
reflexive transitive closure of the interleaved step relation.
\end{definition}

\begin{definition}[\bfseries Semantics of the RG judgment]
\label{def:rg-semantics}
Semantics of the RG sextuple $\rg{\mathbb{I},P,R}{c}{G,Q}$ is defined
in terms of the interleaved step relation as following:
\begin{smathpar}
\begin{array}{l}
\hspace*{-0.3in}
\rg{\mathbb{I},P,R}{c}{G,Q} \;\defeq\; \forall \E.\, P(\E)
  \wedge \mathbb{I}(\E) \\
\hspace*{0.4in}\Rightarrow (\forall n,\E'.\; \I \vdash (c,\E) 
    \rstepssto{n} (\cskip,\E') \Rightarrow Q(\E')) \\
\hspace*{0.5in}\conj \texttt{step-guaranteed}(\I,R,G,c,\E)\\
\end{array}
\end{smathpar}
The first conjunct in the consequent is called the \emph{Hoare
consequent} since it ascribes Hoare triple semantics to RG sextuple.
The second conjunct, called the \emph{guarantee consequent}, uses the
$\texttt{step-guaranteed}$ predicate defined below:
\begin{smathpar}
\begin{array}{l}
\texttt{step-guaranteed}(\I,R,G,c,\E) \;\defeq\; \forall n,\E',c'',\E''.\\
\hspace*{0.2in}\I \vdash (c,\E) \rstepssto{n} (c',\E') \conj \I \vdash (c',\E') \stepsto
  (c'',\E'') \Rightarrow G(\E',\E'')\\
\end{array}
\end{smathpar}
The guarantee consequent requires $G$ to capture the trace effect of
every small-step of $c$, where the reduction can be interleaved by the
interference ($R$) from concurrent threads. Semantics of the RG
sextuple for transaction-bound commands ($\txnbox{c}_i$) is defined
similarly. Expressions, unlike commands, evaluate to a value $v$, and
the semantics of their RG septuple
($\rg{\I,P,R}{\txnbox{e}_i}{G,C,Q}$) differs slightly in that its
Hoare consequent requires the value $v$ to satisfy the assertion $C$. 
\end{definition}

It is worthwhile to note that the semantics of all RG judgments,
including the judgments for transaction-bound terms,  make similar
demands of the guarantee relation. Considering that transactions are
atomic (though not isolated), at a first glance, it might seem odd
that a transaction's guarantee is required to make explicit every step
of its reduction. This requirement is however justifed because, in
reality, a transaction's atomicity is predicated on the isolation
settings of the observer. A \iso{Read Uncommitted} transaction, for
example, can observe the internal state of a transaction, regardless
of the latter's preferences\footnote{Thus, there is no such thing as
atomicity; it's all about isolation.}. In the interest of modular
verification, the transaction must therefore make its internal state
available via its guarantee relation.

Soundness of rely-guarantee judgments can now be stated
thus\footnote{Supplementary contains a rigorous proof of the soundness
theorem, and all relevant lemmas.}:
\begin{theorem}[\bfseries Soundness] 
Rely-guarantee judgments defined by the rules in
Fig.~\ref{fig:rg-rules} are sound with respect to
Definition~\ref{def:rg-semantics}. 
\end{theorem}
\noindent In particular, if $\rg{\I,P,R}{c}{G,Q}$ can be derived using
the rules of Fig.~\ref{fig:rg-rules}, then (a). every interleaved
multi-step reduction of $c$ starting from a trace that satisfies $P$
and $\I$, results in a trace that satisfies $Q$, and (b). the effect
that every small-step of $c$ has on the trace is contained in $G$.
Soundness of RG judgment for transaction-bound commands
($\txnbox{c}_i$) is stated similarly.  For expressions, soundness of
the judgment $\rg{\I,P,R}{\txnbox{e}_i}{G,C,Q}$ also proves that $e$
is always evaluated to a value that satisfies $C$.

>>>>

We now describe a proof system that lets us demonstrate that a \txnimp
program satisfies its high-level invariants in all small-step
executions that satisfy a chosen trace invariant ($\I$).\footnote{It is
important to keep in mind the distinction between high-level invariants
(\eg~\C{X$\ge$0}) and trace invariants (\eg~\C{RC}). The former
constitute \emph{proof} \emph{obligations} for the programmer, whereas
the latter captures \emph{assumptions} about the operational characteristics
of transactions and the underlying data store. } 
Since the trace invariant captures transaction-isolation constraints,
the demonstration is a proof that the given selection of isolation
levels for transactions is sufficient to enforce a program's invariants.

As mentioned earlier, our proof system is a rely-guarantee framework
that admits compositional reasoning by abstracting away 
environmental interference as a rely relation. A conspicuous
difference between a standard development of rely-guarantee and ours
is that, while the former reasons in terms of program states (variable
to value bindings), we reason in terms of executions as captured by
their traces ($\E$). In particular, our rely ($R$) and guarantee ($G$)
relations relate executions (i.e., $R,\,Q\,\subseteq\,\E\times\E$),
and pre- ($P$) and post- ($Q$) conditions as assertions over executions
(i.e., $P,\,Q\, : \E \rightarrow \Prop$). Our development nonetheless
facilitates state-based reasoning via the interpretation function
($\interp{\cdot}$) introduced in \S\ref{sec:syntax}. For example, if a
\emph{bi-state} rely relation relates every pair of states $\sigma$
and $\sigma'$ such that $\sigma'(X) \ge \sigma(X) \ge 0$, the
corresponding \emph{bi-execution} rely relation relates every pair of
executions $\E$ and $\E'$ such that $\interp{\E}(X) \ge
\interp{\E'}(X) \ge 0$.  Assertions on states are also written
similarly. For instance, consider the postcondition
$\C{C\,=\,k-a1-a2}$ of the program in Fig.~\ref{fig:motiv-eg-1}. The
corresponding assertion on the post-state
($\lambda\sigma.~\sigma(\C{C}) \,=\, \C{k-a1-a2}$) asserts that in all
states resulting from executing the program, the value of \C{C} is
\C{k-a1-a2}. The equivalent execution-based assertion
($\lambda\E.~\interp{\E}(\C{C})\,=\,\C{k-a1-a2}$) asserts that in all
executions of the program, the value of \C{C} is \C{k-a1-a2}.
However, having access to an execution facilitates assertions that go
beyond the state. An example is an invariant of \C{Wd1} described
abstractly in \S\ref{sec:motivation}, and reified as an
execution-based assertion below:

\begin{smathpar}
\begin{array}{l}
  \lambda\E.~ \neg(\underE{\committed(\C{Wd1}}) \conj 
      \underE{\C{Wd1} \wrstoar \C{C}}\\
  \hspace*{0.2in}\conj(\neg(\underE{\committed(\C{Wd2})}) \Rightarrow 
      \C{\interp{\E}(C) = k-a1}) \\
   \hspace*{0.2in}\conj (\underE{\committed(\C{Wd2})} \Rightarrow 
      \C{\interp{\E}(C) = k-a1-a2})
\end{array}
\end{smathpar}
% Where, $\committed$ is defined as:
% \begin{smathpar}
% \begin{array}{lcl}
% \underE{\committed(T_i)} & \Leftrightarrow &
%   \exists\eta.~\underE{\eta\in T_i} \conj \kind(\eta) = \C{COMMIT}\\
% \end{array}
% \end{smathpar}

\noindent One of the conjuncts asserts that \C{Wd1} has written to \C{C} - a
fact which cannot be deduced solely from the value of \C{C} (esp. in the
presence of interference), but can be expressed as a proposition over
the execution trace.

@@

The standard rely-guarantee judgment is a quintuple
$\rg{P,R}{c}{G,Q}$, which informally asserts that if a
command $c$ is executed in a state that satisfies its precondition
$P$, provided that every interference step during the execution is
contained inside the rely relation $R$, then the effect that each step
of executing $c$ has on the state is captured by $G$, and the final
state of execution satisfies the postcondition $Q$. The terms
``execution'' and ``step'' are defined with respect to an operational
semantics, which, in our case, is parameterized on the trace invariant
$\I$. Consequently, our rely-guarantee judgment for commands is the
sextuple $\rg{\I,P,R}{c}{G,Q}$, whose semantics differs from the
quintuple in that it requires every step of the execution to additionally preserve
the trace invariant $\I$ for the postcondition to be valid.  As usual,
if $c$ is inside a transaction $T_i$, we write $\txnbox{c}_i$. \txnimp
expressions are side-effecting (they generate \C{RD} effects), and
admit interference during their evaluation.  Therefore, like 
commands, expressions need to be reasoned about explicitly.  In contrast to
commands, however, expressions evaluate to values, and the reasoning
framework should also admit assertions on such values.  We therefore
define a separate judgment for expressions - the septuple
$\rg{\I,P,R}{e}{G,C,Q}$. The new entrant $C : \mathbb{N} \rightarrow
\Prop$ is an assertion on the return value of $e$.

The rules that define rely-guarantee judgment are shown in
Fig.~\ref{fig:rg-rules}. Like standard rely-guarantee definitions,
these definitions also require a \emph{stability} condition, which
requires pre- and post- conditions to hold despite any interference
from concurrent threads (captured by $R$). Stability can be predicated
on the assumption that interference preserves the trace invariant
$\I$. Formally:

\begin{smathpar}
\begin{array}{lcl}
  \underI{\stable(R,P)} & \defeq & \forall \E, \E'.\, 
  \I(\E) \conj P(\E) \conj R(\E,\E') \\
  &   & \hspace*{1in}\conj \I(\E') \Rightarrow P(\E')\\
\end{array}
\end{smathpar}

\noindent However, the assumption that interference preserves $\I$ (or,
dually, $\I$ withstands inteference) needs to be justified
separately. We call this the stability requirement on $\I$:

\begin{smathpar}
\begin{array}{lcl}
\stable(R,\I)& \defeq & \forall \E, \E'.\, 
  \I(\E) \conj R(\E,\E')\Rightarrow \I(\E')\\
\end{array}
\end{smathpar}

The rule \rulelabel{RG-Var} defines the rely-guarantee judgment for
shared variable reads inside a transaction $T_i$. It requires $\I$ to
be stable, and pre- and post- conditions to be stable relative to
$\I$.  The quantified premise effectively requires a proof that if the
abstract machine of Fig.~\ref{fig:txnimp} takes a step starting from
an execution $\E$ that satisfies the precondition $P$, then the
resultant execution $\E'$ satisfies the postcondition $Q$, and that
the guarantee $G$ faithfully captures the transition from $\E$ to
$\E'$. It is informative to compare this premise with the premise of
the \rulelabel{E-Aux} reduction rule of Fig.~\ref{fig:txnimp}. Similar
premises also appear in \rulelabel{RG-Asgn} and \rulelabel{RG-Txn}
rules, which define rely-guarantee judgments for assignments and
transactions, respectively. \rulelabel{RG-Var} however also requires
the return value ($\interp{S}(X)$) of the read to satisfy the
assertion $C$ meant for the value. \rulelabel{RG-Arith} defines the RG
judgment for an arithmetic expression $e_1\pm e_2$ in terms of the
corresponding judgments for the constitutent expressions $e_1$ and
$e_2$. The quantified premise requires any value resulting from
evaluating $e_1 \pm e_2$ to satisfy the assertion $C$, provided that
$e_1$ and $e_2$ always evaluate to values that satisfy $C_1$ and
$C_2$, respectively. The rules for sequential and parallel composition
of commands are essentially the same as their counterparts in a
standard rely-guarantee formulation and hence elided.

@@

The \rulelabel{RG-Conseq} rule define ways to strengthen or weaken
relations and assertions associated with the RG judgment of
transaction-bound expressions. Similar rules exist for
transaction-bound and top-level commands, but are not
shown.\footnote{The supplementary materials provide the complete set
  of rules.} As is the case with a standard rely-guarantee
formulation,the rules allow the precondition $P$ and the rely relation
$R$ to be strengthened\footnote{A stronger relation is a smaller set
  than a weaker relation, hence a stronger $R$ admits fewer
  interferences than a weaker one}, and the postcondition $Q$ (also,
$C$ in the case of expressions) and the guarantee relation $G$ to be
weakened.  The most notable aspect of the \rulelabel{RG-Conseq} rules
is that they allow the trace invariant $\I$ to be
strengthened. Considering that $\I$ captures isolation properties,
this means that a program proven correct under weaker isolation levels
is also correct under stronger ones.  Parametricity over the trace
invariant $\I$, combined with the ability to strengthen $\I$ as
needed, allows our proof system to support a highly flexible proof
strategy to prove programs correct over various isolation
variants. For example, programmers can \emph{define} isolation
guarantees of their choice (by defining $\I$ appropriately) and then
prove programs correct assuming the guarantees hold.  The soundness of
strengthening $\I$ assures that the program can now be executed safely
on any data store that offers isolation guarantees at least as strong
as those assumed.

@@

\begin{definition}[\bfseries Step-indexed reflexive transitive closure]
Forall $A:\text{Type}$, $R: A \rightarrow A \rightarrow \mathbb{P}$, and $n :
\mathbb{N}$, step-indexed reflexive transitive closure $R^n$ of $R$ is
the smallest relation satisfying the following
properties:
\begin{itemize}
\item $\forall (x:A).\, R^0 (x,x)$
\item $\forall (n:\mathbb{N})(x,y,z : A).\, R(x,y) \conj R^n(y,z) \Rightarrow
R^{n+1}(x,z)$
\end{itemize}
\end{definition}

\begin{definition}[\bfseries Interleaved step and multi-step relations]
An Interleaved step relation interleaves thread-local reductions with
interference from concurrent threads captured as the rely relation
($R$).  It is defined thus:
\begin{smathpar}
\begin{array}{lcl}
\I \vdash (c,\E) \rstepsto (c',\E') & \defeq & \I \vdash 
  (c,\E) \stepsto (c',\E') \\
  &   & \disj (c' = c \conj R(\E, \E') \conj \I(\E'))\\
\end{array}
\end{smathpar}
The interleaved step relation for transaction bound expressions
($\txnbox{e}_i$) and commands ($\txnbox{c}_i$) is defined similarly.
An Interleaved multi-step relation ($\stepssto{n}$) is the step-indexed
reflexive transitive closure of the interleaved step relation.
\end{definition}

\begin{definition}[\bfseries Semantics of the RG judgment]
\label{def:rg-semantics}
The semantics of the RG sextuple $\rg{\mathbb{I},P,R}{c}{G,Q}$ is defined
in terms of the interleaved step relation thus:
\begin{smathpar}
\begin{array}{l}
\hspace*{-0.3in}
\rg{\mathbb{I},P,R}{c}{G,Q} \;\defeq\; \forall \E.\, P(\E)
  \wedge \mathbb{I}(\E) \\
\hspace*{0.4in}\Rightarrow (\forall n,\E'.\; \I \vdash (c,\E) 
    \rstepssto{n} (\cskip,\E') \Rightarrow Q(\E')) \\
\hspace*{0.5in}\conj \texttt{step-guaranteed}(\I,R,G,c,\E)\\
\end{array}
\end{smathpar}
The first conjunct in the consequent is called the \emph{Hoare
consequent} since it ascribes Hoare triple semantics to an RG sextuple.
The second conjunct, called the \emph{guarantee consequent}, uses the
$\texttt{step-guaranteed}$ predicate defined below:
\begin{smathpar}
\begin{array}{l}
\texttt{step-guaranteed}(\I,R,G,c,\E) \;\defeq\; \forall n,\E',c'',\E''.\\
\hspace*{0.2in}\I \vdash (c,\E) \rstepssto{n} (c',\E') \conj \I \vdash (c',\E') \stepsto
  (c'',\E'') \Rightarrow G(\E',\E'')\\
\end{array}
\end{smathpar}
The guarantee consequent requires $G$ to capture the trace effect of
every small-step of $c$, where the reduction can be interleaved by the
interference ($R$) from concurrent threads. The semantics of the RG
sextuple for transaction-bound commands ($\txnbox{c}_i$) is defined
similarly. Expressions, unlike commands, evaluate to a value $v$, and
the semantics of their RG septuple ($\rg{\I,P,R}{\txnbox{e}_i}{G,C,Q}$) differs slightly in that its
Hoare consequent requires the value $v$ to satisfy the assertion $C$. 
\end{definition}

Note that the semantics of all RG judgments, including the judgments
for transaction-bound terms, make similar demands of the guarantee
relation. Given that transactions are atomic (though not isolated), it
is not immediately apparent why a transaction's guarantee is required
to make explicit every step of its reduction. This requirement is
justifed however because, in reality, a transaction's atomicity is
predicated on the isolation settings of the observer. A \iso{Read
  Uncommitted} transaction, for example, is permitted to observe the
internal state of a transaction $T$ even if $T$ is claimed to execute
atomically.  In the interest of modular verification, the transaction
must therefore make its internal state available via its guarantee
relation.

\noindent The soundness of rely-guarantee judgments can be stated
thus\footnote{Full proofs are provided in the supplementary material.}:
\begin{theorem}[\bfseries Soundness] 
The rely-guarantee judgments defined by the rules in
Fig.~\ref{fig:rg-rules} are sound with respect to
Definition~\ref{def:rg-semantics}. 
\end{theorem}
\noindent In particular, if $\rg{\I,P,R}{c}{G,Q}$ can be derived using
the rules of Fig.~\ref{fig:rg-rules}, then (a) every interleaved
multi-step reduction of $c$ starting from a trace that satisfies $P$
and $\I$, results in a trace that satisfies $Q$, and (b) the effect
that every small-step of $c$ has on the trace is contained in $G$.
Soundness of the RG judgment for transaction-bound commands
($\txnbox{c}_i$) is stated similarly.  For expressions, soundness of
the judgment $\rg{\I,P,R}{\txnbox{e}_i}{G,C,Q}$ also proves that $e$
is always evaluated to a value that satisfies $C$.
